---
title: "STA6206 - Bayesian Data Analysis - Practical 3 (Solutions)"
author: "Marc Henrion"
date: "11 September 2019"
header-includes:
   - \usepackage{amsmath}
output:
  powerpoint_presentation:
    reference_doc: MlwChanco_Template.pptx
---

```{r setup, include=FALSE, echo=F}
knitr::opts_chunk$set(echo = TRUE, fig.width=16, fig.height=9, dpi=150, highlight=T)

require(tidyverse)
require(knitr)
require(gridExtra)
```


#

## Practical 3

# Notation

* $X, Y, Z$ - random variables

* $x, y, z$ - measured / observed values

* $\bar{X}$, $\bar{Y}, \bar{Z}$ - sample mean estimators for X, Y, Z

* $\bar{x}$, $\bar{y}, \bar{z}$ - sample mean estimates of X, Y, Z

* $\hat{T}$, $\hat{t}$ - given a statistic T, estimator and estimate of T

* $P(A)$ - probability of an event A occuring

* $f_X(.)$, $f_Y(.), f_Z(.)$ - probability mass / density functions of X, Y, Z; sometimes $p_X(.)$ etc. rather than $f_X(.)$

* p(.) - used as a shorthand notation for pmfs / pdfs if the use of this is unambiguous (i.e. it is clear which is the random variable) 

* $X\sim F$ - X distributed according to distribution function F

* $E[X]$, $E[Y]$, $E[Z]$, $E[T]$ - the expectation of X, Y, Z, T respectively



# Exercise 1

Show that the Bayes estimator $\hat{\theta}_B$ for the quadratic loss function $\mathcal{C}(\theta-\hat{\theta})=(\theta-\hat{\theta})^2$ is given by the posterior mean. In other words, show that:

$$\,$$

$$E[\theta|y]=\arg\min_{\hat{\theta}}\int_\mathcal{Y}\int_\Theta\mathcal{C}(\theta-\hat{\theta})p(\theta,y)d\theta dy$$


# Exercise 1 (Solution)

We saw in lectures that we can use the multiplication rule and then optimise the inner integral only:

$$
\begin{align}
\hat{\theta}_B &=& \arg\min_{\hat{\theta}}\int \mathcal{C}(\theta-\hat{\theta})p(\theta|y)d\theta \\
               &=& \arg\min_{\hat{\theta}}\int (\theta-\hat{\theta})^2p(\theta|y)d\theta
\end{align}
$$

To find the minimum, we solve

$$
\begin{align}
     & \frac{d}{d\hat{\theta}}\int (\theta-\hat{\theta})^2p(\theta|y)d\theta = 0 \\
\iff & \int2(\theta-\hat{\theta})(-1)p(\theta|y)d\theta=0 \\
\iff & \int(\theta-\hat{\theta})p(\theta|y)d\theta=0 \\
\iff &\int\theta p(\theta|y)d\theta=\int\hat{\theta} p(\theta|y)d\theta
\end{align}
$$

# Exercise 1 (Solution)

Note that $\int\hat{\theta} p(\theta|y)d\theta = \hat{\theta} \int p(\theta|y)d\theta=\hat{\theta}$ since the posterior distribution for $\theta$ is a probability distribution and needs to integrate to 1.

$$\,$$

Therefore we see that

$$\,$$

$$
\hat{\theta}_B=\int\theta p(\theta|y) d\theta
$$

$$\,$$

This is the posterior mean $E[\theta|y]=\int\theta p(\theta|y) d\theta$.


# Exercise 2

Suppose $\pi\sim\mbox{Beta}(2,3)$ and $Y_1,\ldots,Y_n\sim_{\mbox{iid}}Bernoulli(\pi)$. Further suppose we observe data $y_1,\ldots,y_n$ with $n=25, k=\sum_i y_i= 16$.

Find the following:

* posterior distribution $p(\pi|k)$ and plot it, comparing it to the prior distribution

* posterior predictive distribution $p(\tilde{y}|y_1,\ldots,y_n)$

* the 95% quantile-based Bayesian confidence interval for $\pi$

* the 95% HPD interval

Further, compute:

* $P(\pi>0.5|k)$

* For the following 2 hypotheses: $H_1:\pi\in[0.3,0.5], H_2:\pi\in[0.5,0.7]$, compute the prior and posterior odds and calculate the Bayes factor.


# Exercise 2 (Solution)

Posterior distribution

$$p(\pi|k=16)\propto\pi^{2-1}(1-\pi)^{3-1}\pi^{16}(1-\pi)^{25-16}=\pi^{17}(1-\pi)^{11}$$
$$\Rightarrow \Pi|k\sim\mbox{Beta}(18,12)$$

```{r}
xx<-seq(0,1,length=100)
yyPr<-dbeta(xx,2,3)
yyPo<-dbeta(xx,18,12)

plot(type="l",xx,yyPo,lwd=3,col="mediumorchid",xlab=expression(pi),ylab="density",cex.lab=2,cex.axis=2)
lines(xx,yyPr,lwd=3,lty=2,col="steelblue")

legend(x="topleft",bty="n",lwd=2,col=c("steelblue","mediumorchid"),lty=c(2,1),legend=c("prior","posterior"),cex=2)
```


# Exercise 2 (Solution)

Posterior predictive distribution

$$\,$$

$$
\begin{align}
p(\tilde{Y}=1|y_1,\ldots,y_n) &=& \int p(\tilde{y},\pi|y_1,\ldots,y_n)d\pi\\
                              &=& \int p(\tilde{y}|pi,y_1,\ldots,y_n)p(\pi|y_1,\ldots,y_n)d\pi \\
                              &=& \int \pi \frac{\Gamma(30)}{\Gamma(18)\Gamma(12)}\pi^{17}(1-\pi)^{11} d\pi \\
                              &=& \frac{\Gamma(30)\Gamma(19)}{\Gamma(18)\Gamma(31)}\int \frac{\Gamma(31)}{\Gamma(19)\Gamma(12)} \pi^{18}(1-\pi)^{11}d\pi \\
                              &=& \frac{\Gamma(30)\Gamma(19)}{\Gamma(18)\Gamma(31)} = \frac{18}{30} = 3/5 = 0.6
\end{align}
$$

From this it follows that $p(\tilde{Y}=0|y_1,\ldots,y_n)=1-\frac{3}{5}=\frac{2}{5}=0.4$.


# Exercise 2 (Solution)

We have:

* $q_{0.025;\mbox{Beta}(18,12)}=$ `r round(digits=4,qbeta(0.025,18,12))`

* $q_{0.975;\mbox{Beta}(18,12)}=$ `r round(digits=4,qbeta(0.975,18,12))`

$$\,$$

Therefore the 95% quantile based Bayesian confidence interval for $\pi$ is given by $[$```r round(digits=2,qbeta(0.025,18,12))```,```r round(digits=2,qbeta(0.975,18,12))```$]$


```{r}
qbeta(c(0.025,0.975),18,12)
```

# Exercise 2 (Solution)

For the HPD we are going to use the function `hdi()` from the R package `HDInterval`:

```{r}
library(HDInterval)
hdp<-hdi(qbeta,0.95,shape1=18,shape2=12)
print(hdp)
```

From this we find that the 95% HDP interval for $\pi$ is given by $[$```r round(digits=2,hdp[1])```,```r round(digits=2,hdp[2])```$]$


# Exercise 2 (Solution)

We have

$$
\begin{align}
P(\pi>0.5|Y=k) &=& \int_{0.5}^1p(\pi|k)d\pi \\
               &=& \int_{0.5}^1\frac{\Gamma(30)}{\Gamma(18)\Gamma(12)}\pi^{17}(1-\pi)^{11}d\pi \\
               &=& \approx0.87
\end{align}
$$

```{r}
integrate(dbeta,lower=0.5,upper=1,shape1=18,shape2=12)
```

Alternatively we could have worked with the posterior cdf:

```{r}
1-pbeta(0.5,18,12)
```

# Exercise 2 (Solution)

Prior odds

$$\frac{P(H_1)}{P(H_2)}=\frac{\int_{0.3}^{0.5}p_{\beta(3,2)}(\pi)d\pi}{\int_{0.5}^{0.7}p_{\beta(3,2)}(\pi)d\pi}=1.48$$

$$\,$$

Prosterior odds

$$\frac{P(H_1|k)}{P(H_2|k)}=\frac{\int_{0.3}^{0.5}p_{\beta(18,12)}(\pi)d\pi}{\int_{0.5}^{0.7}p_{\beta(18,12)}(\pi)d\pi}=0.18$$

# Exercise 2 (Solution)

```{r}
priorOdds<-integrate(dbeta,0.3,0.5,shape1=2,shape2=3)$value/integrate(dbeta,0.5,0.7,shape1=2,shape2=3)$value
#priorOdds<-(pbeta(0.5,2,3)-pbeta(0.3,2,3))/(pbeta(0.7,2,3)-pbeta(0.5,2,3)) # alternative way of doing this
priorOdds
posteriorOdds<-integrate(dbeta,0.3,0.5,shape1=18,shape2=12)$value/integrate(dbeta,0.5,0.7,shape1=18,shape2=12)$value
posteriorOdds
```


# Exercise 2 (Solution)

Bayes Factor

$\mbox{BF} = \frac{\mbox{posterior odds}}{\mbox{prior odds}}=0.18/1.48 = 0.12$

$$\,$$

```{r}
bayesFactor<-posteriorOdds/priorOdds
bayesFactor
```


# Exercise 3

Suppose $\pi\sim\mbox{Gamma}(5,2)$ and $Y_1,\ldots,Y_n\sim_{\mbox{iid}}Poisson(\lambda)$. Further suppose we observe data $y_1,\ldots,y_n$ with $n=18, k=\sum_i y_i= 40$.

Find the following:

* posterior distribution $p(\lambda|y_1,\ldots,y_n$ and plot it, comparing it to the prior distribution

* posterior predictive distribution $p(\tilde{y}|y_1,\ldots,y_n)$

* the 95% quantile-based Bayesian confidence interval for $\lambda$

* the 95% HPD interval

Further, compute:

* $P(\lambda\leq1|y_1,\ldots,n)$

* For the following 2 hypotheses: $H_1:\lambda\in[0.75,1.25], H_2:\lambda\in[1.75,2.25]$, compute the prior and posterior odds and calculate the Bayes factor.


# Exercise 3 (Solution)

Posterior distribution

$$
\begin{align}
p(\lambda|y_1,\ldots,y_n) &\propto& \lambda^{5-1}e^{-2\lambda}\lambda^{\sum_i y_i} e^{-n\lambda} \\
                          &\propto& \lambda^{5+40-1}e^{-(2+18)\lambda} \\
\Rightarrow \Lambda|y_1,\ldots,y_n &\sim& \Gamma(45,20)
\end{align}
$$

```{r}
xx<-seq(0,8,length=100)
yyPr<-dgamma(xx,shape=5,rate=2)
yyPo<-dgamma(xx,shape=45,rate=20)

plot(type="l",xx,yyPo,lwd=3,col="mediumorchid",xlab=expression(pi),ylab="density",cex.lab=2,cex.axis=2)
lines(xx,yyPr,lwd=3,lty=2,col="steelblue")

legend(x="topleft",bty="n",lwd=2,col=c("steelblue","mediumorchid"),lty=c(2,1),legend=c("prior","posterior"),cex=2)
```


# Exercise 3 (Solution)

Posterior predictive distribution

$$\,$$

From Practical 1&2, Exercise 4, we know:

$$\,$$

$$\tilde{Y}|y_1,\ldots,y_n\sim\mbox{NegBin}(45,20/21)$$


# Exercise 3 (Solution)

We have:

* $q_{0.025;\Gamma(45,20)}=$ `r round(digits=4,qgamma(0.025,shape=45,rate=20))`

* $q_{0.975;\Gamma(45,20)}=$ `r round(digits=4,qgamma(0.975,shape=45,rate=20))`

$$\,$$

Therefore the 95% quantile based Bayesian confidence interval for $\pi$ is given by $[$```r round(digits=2,qgamma(0.025,shape=45,rate=20))```,```r round(digits=2,qgamma(0.975,shape=45,rate=20))```$]$

```{r}
qgamma(c(0.025,0.975),shape=45,rate=20)
```

# Exercise 3 (Solution)

For the HPD we are going to use the function `hdi()` from the R package `HDInterval`:

```{r}
library(HDInterval)
hdp<-hdi(qgamma,0.95,shape=45,rate=20)
print(hdp)
```

From this we find that the 95% HDP interval for $\lambda$ is given by $[$```r round(digits=2,hdp[1])```,```r round(digits=2,hdp[2])```$]$


# Exercise 3 (Solution)

We have

$$
\begin{align}
P(\lambda\leq1|y_1,\ldots,y_n) &=& \int_{0}^1p(\lambda|y_1,\ldots,y_n)d\lambda \\
               &=& \int_0^1 \frac{20^{45}}{\Gamma(45)}\lambda^{44}e^{-20\lambda}d\lambda \\
               &=& 1.06\cdot10^{-6}
\end{align}
$$

```{r}
integrate(dgamma,lower=0,upper=1,shape=45,rate=20)
```


# Exercise 3 (Solution)


Prior odds

$$\frac{P(H_1)}{P(H_2)}=\frac{\int_{0.75}^{1.25}p_{\gamma(5,2)}(\lambda)d\lambda}{\int_{1.75}^{2.25}p_{\Gamma(5,2)}(\lambda)d\lambda}=0.2513$$

$$\,$$

Prosterior odds

$$\frac{P(H_1|y_1,\ldots,y_n)}{P(H_2|y_1,\ldots,y_n)}=\frac{\int_{0.75}^{1.25}p_{\gamma(45,20)}(\lambda)d\lambda}{\int_{1.75}^{2.25}p_{\Gamma(45,20)}(\lambda)d\lambda}=0.0004$$

# Exercise 3 (Solution)

```{r}
priorOdds<-integrate(dgamma,0.75,1.25,shape=5,rate=2)$value/integrate(dgamma,1.25,2.25,shape=5,rate=2)$value
priorOdds
posteriorOdds<-integrate(dgamma,0.75,1.25,shape=45,rate=20)$value/integrate(dgamma,1.25,2.25,shape=45,rate=20)$value
posteriorOdds
```


# Exercise 3 (Solution)

Bayes Factor

$\mbox{BF} = \frac{\mbox{posterior odds}}{\mbox{prior odds}}=0.004/0.2513 = 0.0015$

$$\,$$

```{r}
bayesFactor<-posteriorOdds/priorOdds
bayesFactor
```



#

[end of STA6206 BDA Practical 3]
