---
title: "STA6206 - Bayesian Data Analysis - Practical 5 (Solutions)"
author: "Marc Henrion"
date: "13 September 2019"
header-includes:
   - \usepackage{amsmath}
output:
  powerpoint_presentation:
    reference_doc: MlwChanco_Template.pptx
---

```{r setup, include=FALSE, echo=F}
knitr::opts_chunk$set(echo = TRUE, fig.width=16, fig.height=9, dpi=150, highlight=T)

require(tidyverse)
require(knitr)
require(gridExtra)
```


#

## Practical 5

# Notation

* $X, Y, Z$ - random variables

* $x, y, z$ - measured / observed values

* $\bar{X}$, $\bar{Y}, \bar{Z}$ - sample mean estimators for X, Y, Z

* $\bar{x}$, $\bar{y}, \bar{z}$ - sample mean estimates of X, Y, Z

* $\hat{T}$, $\hat{t}$ - given a statistic T, estimator and estimate of T

* $P(A)$ - probability of an event A occuring

* $f_X(.)$, $f_Y(.), f_Z(.)$ - probability mass / density functions of X, Y, Z; sometimes $p_X(.)$ etc. rather than $f_X(.)$

* p(.) - used as a shorthand notation for pmfs / pdfs if the use of this is unambiguous (i.e. it is clear which is the random variable) 

* $X\sim F$ - X distributed according to distribution function F

* $E[X]$, $E[Y]$, $E[Z]$, $E[T]$ - the expectation of X, Y, Z, T respectively



# Exercise 1

$$\,$$

Fit the model from Practical 3, Exercise 3 using JAGS and the `rjags` package.
Use this as the data from the sampling model:
$$y=(1, 3, 2, 3, 0, 2, 6, 4, 4, 1, 1, 3, 2, 3, 1, 1, 3, 0)$$

$$\,$$

Inspect the trace plot and plot the posterior distribution.

$$\,$$

Compute the posterior mean and the quantile-based 95% Bayesian confidence interval.


# Exercise 1 (Solution)

Write the following JAGS model into a file called `jagsP5ex1.jags`:

    model{
      # sampling model
      for(i in 1:N){
        y[i]~dpois(lambda)
      }
      
      # prior
      lambda~dgamma(5,2)
    }
    
This specifies the model. Now we need to fit this model using MCMC.

For this we use `R` and the `rjags` library.


# Exercise 1 (Solution)

$$\,$$

```{r, collapse=T}
library(rjags)

set.seed(123)

dat<-list(N=18,y=c(1,3,2,3,0,2,6,4,4,1,1,3,2,3,1,1,3,0)) # 18 observations, y_i sum to 40

jagsMod<-jags.model("jagsP5ex1.jags",data=dat,n.chains=4,n.adapt=1000)
```

# Exercise 1 (Solution)

$$\,$$

```{r, collapse=T}
update(jagsMod,1000)
parsPosterior<-coda.samples(model=jagsMod,variable.names=c("lambda"),n.iter=1e4)

# check trace plot and empirical posterior distribution
plot(parsPosterior)

# posterior mean estimate
summary(parsPosterior)$statistics["Mean"]

# posterior quantile based 95% credible interval
summary(parsPosterior)$quantiles[c("2.5%","97.5%")]
```


# Exercise 2

Generate the following data

```{r, collapse=T}
N<-100
x<-rnorm(N)
z<-2-4*x
p<-1/(1+exp(-z))
y<-rbinom(n=N,size=1,prob=p)

dat<-list(N=N,x=x,y=y)
```

Use `R` and JAGS to fit a Bayesian logistic regression model to this data:

$$g(E[Y|X])=\beta_0+\beta_1X$$
where $g(\pi)=\log(\pi/(1-\pi))$


# Exercise 2 (Solution)

JAGS model file (save as `jagsP5ex2.jags`):

    model{
      # logistic regression model
      for(i in 1:N){
        y[i]~dbern(p[i])
        p[i]<-1/(1+exp(-z[i]))
        z[i]<-b0+b1*x[i]
      }
      
      # priors
      b0~dnorm(0,0.0001)
      b1~dnorm(0,0.0001)
    }

# Exercise 2 (Solution)

```{r, collapse=T}
library(rjags)

set.seed(123)

jagsMod<-jags.model("jagsP5ex2.jags",data=dat,n.chains=4,n.adapt=1000)
```

```{r, collapse=T}
update(jagsMod,1000)
parsPosterior<-coda.samples(model=jagsMod,variable.names=c("b0","b1"),n.iter=1e4)

# check trace plot and empirical posterior distribution
plot(parsPosterior)

# posterior mean estimate
summary(parsPosterior)$statistics[,"Mean"]

# posterior quantile based 95% credible interval
summary(parsPosterior)$quantiles[,c("2.5%","97.5%")]
```

#

[end of STA6206 BDA Practical 5]
