---
title: "STA6206 - Bayesian Data Analysis - Practical 4"
author: "Marc Henrion"
date: "12 September 2019"
header-includes:
   - \usepackage{amsmath}
output:
  powerpoint_presentation:
    reference_doc: MlwChanco_Template.pptx
---

```{r setup, include=FALSE, echo=F}
knitr::opts_chunk$set(echo = TRUE, fig.width=16, fig.height=9, dpi=150, highlight=T)

require(tidyverse)
require(knitr)
require(gridExtra)
```


#

## Practical 4

# Notation

* $X, Y, Z$ - random variables

* $x, y, z$ - measured / observed values

* $\bar{X}$, $\bar{Y}, \bar{Z}$ - sample mean estimators for X, Y, Z

* $\bar{x}$, $\bar{y}, \bar{z}$ - sample mean estimates of X, Y, Z

* $\hat{T}$, $\hat{t}$ - given a statistic T, estimator and estimate of T

* $P(A)$ - probability of an event A occuring

* $f_X(.)$, $f_Y(.), f_Z(.)$ - probability mass / density functions of X, Y, Z; sometimes $p_X(.)$ etc. rather than $f_X(.)$

* p(.) - used as a shorthand notation for pmfs / pdfs if the use of this is unambiguous (i.e. it is clear which is the random variable) 

* $X\sim F$ - X distributed according to distribution function F

* $E[X]$, $E[Y]$, $E[Z]$, $E[T]$ - the expectation of X, Y, Z, T respectively



# Exercise 1

Let's revisit Exercise 5 from Practical 1&2.

We had 2 groups of women and we compared the number of children born to each women in the 2 groups. For each group we assumed a Poisson sampling model: $Y_{i,j}\sim\mbox{Pois}(\theta_i), i=1,\ldots,n_j, j=1,2$ and we found that the posterior distributions were:

1. Women without college degree: $\theta_1\sim\Gamma(219,112)$

2. Women with college degree: $\theta_2\sim\Gamma(68,45)$

We had computed $P(\theta_1>\theta_2|n_1,n_2,\sum_iy_{i,1},\sum_i y_{i,2})=0.97$.

Use the Monte Carlo method to compute

$$P(\tilde{Y}_1>\tilde{Y}_2|n_1,n_2,\sum_iy_{i,1},\sum_i y_{i,2})$$


# Exercise 1 (cont'd)

For the group of women without college degree, remember that we found that the posterior predictive distribution was a negative binomial:

$$
\tilde{Y}_1|n_1,\sum_i y_{i,1}\sim\mbox{NegBin}(219,112/113)
$$

Compare this distribution with the empirical distribution of the raw data (given on next slide).

Let $\mathbf{y}=(y_{1,1},\ldots,y_{n_1,1})$. Define $t(\mathbf{y})$ as the ratio of 2's in $\mathbf{y}$ to the number of $1's$. In this dataset we observe $t(\mathbf{y})=38/19=2$. Use the posterior predictive distribution for $\tilde{Y}_1|n_1,\sum_i y_{i,1}$ and the Monte Carlo method to compute $P(t(\mathbf{Y})\geq2)$. What is your conclusion?


# Exercise 1 (cont'd)

```{r, echo=F}
countMat<-cbind(0:6,c(20,19,38,20,10,2,2))
colnames(countMat)<-c("no. children per mother","number of mothers")

kable(countMat)
```


#

[end of STA6206 BDA Practical 4]
