---
title: "STA6206 - Bayesian Data Analysis - Practical 5 (Solutions)"
author: "Marc Henrion"
date: "13 September 2019"
header-includes:
   - \usepackage{amsmath}
output:
  powerpoint_presentation:
    reference_doc: MlwChanco_Template.pptx
---

```{r setup, include=FALSE, echo=F}
knitr::opts_chunk$set(echo = TRUE, fig.width=16, fig.height=9, dpi=150, highlight=T)

require(tidyverse)
require(knitr)
require(gridExtra)
```


#

## Practical 5

# Notation

* $X, Y, Z$ - random variables

* $x, y, z$ - measured / observed values

* $\bar{X}$, $\bar{Y}, \bar{Z}$ - sample mean estimators for X, Y, Z

* $\bar{x}$, $\bar{y}, \bar{z}$ - sample mean estimates of X, Y, Z

* $\hat{T}$, $\hat{t}$ - given a statistic T, estimator and estimate of T

* $P(A)$ - probability of an event A occuring

* $f_X(.)$, $f_Y(.), f_Z(.)$ - probability mass / density functions of X, Y, Z; sometimes $p_X(.)$ etc. rather than $f_X(.)$

* p(.) - used as a shorthand notation for pmfs / pdfs if the use of this is unambiguous (i.e. it is clear which is the random variable) 

* $X\sim F$ - X distributed according to distribution function F

* $E[X]$, $E[Y]$, $E[Z]$, $E[T]$ - the expectation of X, Y, Z, T respectively



# Exercise 1

$$\,$$

Fit the model from Practical 3, Exercise 3 using JAGS and the `rjags` package.
Use this as the data from the sampling model:
$$y=(1, 3, 2, 3, 0, 2, 6, 4, 4, 1, 1, 3, 2, 3, 1, 1, 3, 0)$$

$$\,$$

Inspect the trace plot and plot the posterior distribution.

$$\,$$

Compute the posterior mean and the quantile-based 95% Bayesian confidence interval.


# Exercise 2

Generate the following data

```{r, collapse=T}
N<-100
x<-rnorm(N)
z<-2-4*x
p<-1/(1+exp(-z))
y<-rbinom(n=N,size=1,prob=p)

dat<-list(N=N,x=x,y=y)
```

Use `R` and JAGS to fit a Bayesian logistic regression model to this data:

$$g(E[Y|X])=\beta_0+\beta_1X$$
where $g(\pi)=\log(\pi/(1-\pi))$

#

[end of STA6206 BDA Practical 5]
